---
title: "R Notebook"
output: html_notebook
---
# Load required libraries
```{r}
# Load required libraries

## organizational functions
library(tidyr)
library(dplyr)
library(stringr)
library(lubridate)
library(kableExtra)
library(data.table)
library(plyr)                                                 
library(ggh4x)
library(reshape)

# used during the second lollipop graph prep
library(forcats)

# improve the import functions for data within repositories
library(readr)
library(here)

# graphing functions
library(ggplot2)
library(patchwork) # required for "plot layout" when merging lollipop graphs

# packages used in summary data sets
library(flextable)
library(officer)
library(purrr)

# package for running statistical tests
library(car)

# for the Bayesian summaries
library(AER)
library(vcd)
library(skellam)
```


# Import and prepare dataset
```{r} 
################################################    
# import and prepare a combined dataset
################################################ 

# read the appropriate CSV from GitHub into R
all_pc_aru_IMPORT <- readr::read_csv(here::here("CSVs/analysis", "all_occupancy_analysis.csv"))

# calculate richness for ARU and PC separately
pc_aru_richness <- all_pc_aru_IMPORT %>%
 filter(!CommonName %in% c("----", "Trumpeter Swan", "Sandhill Crane", "Snowy Egret", "Forster's Tern", "Sedge Wren", "Common Tern", "Wilson's Snipe")) %>% # had to add the sedge wren and common tern
  dplyr::filter(IndivCount >=1) %>%
  dplyr::group_by(Protocol, Route, Year, Point) %>%
  dplyr::summarize(Richness = length(unique(BirdCd)))

# calculate richness when ARU + PC are added together
pc_aru_richness_added_together <- all_pc_aru_IMPORT %>%
  filter(!CommonName %in% c("----", "Trumpeter Swan", "Sandhill Crane", "Snowy Egret", "Forster's Tern", "Sedge Wren", "Common Tern", "Wilson's Snipe")) %>%
  dplyr::filter(IndivCount >=1) %>%
  dplyr::group_by(Route, Point, Year) %>% # exclude Protocol to get combined richness
  dplyr::summarize(Richness = length(unique(BirdCd))) %>%
  dplyr::mutate(Protocol = "ARU + PC")

# combine the resulting DFs into one DF that has 3 protocol (ARU alone, PC alone, ARU + PC combined)
all_richness_no0 <- rbind(pc_aru_richness, pc_aru_richness_added_together)

# zero-fill richness summary  
all_data_plot <- all_richness_no0 %>%
  dplyr::select(Point, Protocol, Richness, Route, Year) %>%
  pivot_wider(names_from = Protocol, values_from = Richness) %>%
  pivot_longer(cols = c(ARU, PC, `ARU + PC`), names_to = "Protocol", values_to = "Richness") %>%
  dplyr::mutate(Richness = ifelse(is.na(Richness), 0, Richness)) 
```

# Global comparisons of ARU v. PC richness
```{r}
################################################ 
# 4. Richness Summaries in Tables and Graphs
################################################ 

# Visual analysis: what are the distributions of the data? Check with density plots ==============================================================================================
# Compute IQR limits for each Protocol level
iqr_df <- all_data_plot %>%
  dplyr::group_by(Protocol) %>%
  dplyr::summarise(
    q1 = quantile(Richness, 0.25, na.rm = TRUE),
    q3 = quantile(Richness, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

# Compute density curve values for each Protocol level
density_df <- all_data_plot %>%
  filter(!is.na(Richness), !is.na(Protocol)) %>%
  group_by(Protocol) %>%
  reframe({
    d <- density(Richness, na.rm = TRUE)
    data.frame(Richness = d$x, density = d$y)
  }) %>%
  left_join(iqr_df, by = "Protocol") %>%
  filter(Richness >= q1, Richness <= q3)

# define graph aesthetics (same colors as larger plot); this code is redundant
    Protocol_colors <- c(
      "ARU" = "#F8766D",
      "PC"  = "#00BFC4",
      "ARU + PC" = "#6F3959"
    )

# create plot ==============================================================================================
RichDensity <- ggplot(all_data_plot, aes(x = Richness, fill = Protocol)) +
  scale_fill_manual(values = Protocol_colors) +
  geom_density(alpha = 0.3) + # this alpha controls overall shading
  geom_area(
    data = density_df,
    aes(x = Richness, y = density, fill = Protocol, group = Protocol),
    alpha = 0.6, # this one controls IQR shading
    position = "identity",
    inherit.aes = FALSE) +
  theme_classic()

# run plot
RichDensity

# general summary statistics ==============================================================================================
# General richness comparison statistics
richnessSummary <- ddply(all_data_plot, c("Protocol"),
                 summarise,
                 dataRep = sum(!is.na(Richness)),
                 dataMean = mean(Richness, na.rm = T),
                 dataSD = sd(Richness, na.rm = T),
                 dataSE = dataSD / sqrt(dataRep),
                 dataCI95 = dataSE * 1.96)
# ARU: avg richness is 6.20 ± 0.26 SE
# Combined: 6.60 ± 0.27 SE
# PC: 3.12 ± 0.40 SE

# create a file with these summary statistics for global results
write.csv(richnessSummary, here::here("results", "richness_summary_statistics.csv"))



```

# Lollipop graph: ARUs vs. point counts, total detections (with sites labeled)
```{r}
################################################ 
# Create Lollipop Plot!
################################################ 
# Step 1: separate the DF into two halves ==============================================================================================
routes <- sort(unique(all_data_plot$Route))

half <- ceiling(length(routes) / 2)

routes_1 <- routes[1:half]
routes_2 <- routes[(half + 1):length(routes)]

data_1 <- all_data_plot %>% dplyr::filter(Route %in% routes_1)
data_2 <- all_data_plot %>% dplyr::filter(Route %in% routes_2)

# Step 2: create the graph for the first half of the data ==============================================================================================

# Step 2.1: create a DF for the segment lines 

segment_lines1 <- data_1 %>%
  dplyr::arrange(Route, Year, Point, Richness) %>%
  dplyr::group_by(Route, Year, Point) %>%
  dplyr::mutate(next_richness = lead(Richness)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(next_richness))


# Step 2.2: define graph aesthetics 
Protocol_colors <- c(
  "ARU" = "#F8766D",
  "PC"  = "#00BFC4",
  "ARU + PC" = "#6F3959"
)

Protocol_shapes <- c(
  "ARU" = 16,
  "PC"  = 16,
  "ARU + PC" = 18 # diamond
)

# Step 2.3: build the graph

# creation of the plot
richness_plot1 <- ggplot(data_1, # requires long-form data
                         aes(x = Richness, y = Point,
                             color = Protocol, shape = Protocol)) +
  
  # add in lines for each location
  geom_segment(
    data = segment_lines1,
    aes(x = Richness, xend = next_richness,
        y = Point, yend = Point),
    inherit.aes = FALSE,
    color = "gray60",
    linewidth = 1
  ) +
  
  # add in all data points
  geom_point(aes(x = Richness),
             size = 5) +
  # group the data by site (y) and by year (x)
  ggh4x::facet_nested(
    Route ~ Year,
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  
  # reorder the legend and give the points custom colors and shapes
  scale_color_manual(
    values = Protocol_colors,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count combined")
  ) +
  scale_shape_manual(
    values = Protocol_shapes,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count combined")
  ) +
  
  # theme adjustments
  theme_minimal(base_size = 14) + # can change this back to 16 after more troubleshooting
  theme(
    strip.text.y.left = element_text(angle = 0, face = "bold"), # makes Route labels horizontal
    strip.placement = "outside",
    strip.text = element_text(face = "bold"), # makes site, year labels bold
    axis.text.y = element_blank(), # first line to remove the individual Point labels
    axis.ticks.y = element_blank(), # second line to remove the individual Point labels
    panel.spacing = unit(1.2, "lines"),
    panel.background = element_rect(fill = "grey97", color = NA), # add in dark background within each site/year
    legend.position = "top" # if you want the legend at the top, otherwise it'll default to the right.
  ) +
  labs(y = NULL, x = "Species Detected") # remove "Point" from appearing on the left

# Step 2.4: Run the graph to inspect for any errors
#richness_plot1


# Step 3: create the graph for the second half of the data ==============================================================================================

# Step 3.1: create a DF for the segment lines 
segment_lines2 <- data_2 %>%
  dplyr::arrange(Route, Year, Point, Richness) %>%
  dplyr::group_by(Route, Year, Point) %>%
  dplyr::mutate(next_richness = lead(Richness)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(next_richness))


# Step 3.2: define graph aesthetics 
Protocol_colors <- c(
  "ARU" = "#F8766D",
  "PC"  = "#00BFC4",
  "ARU + PC" = "#6F3959"
)

Protocol_shapes <- c(
  "ARU" = 16,
  "PC"  = 16,
  "ARU + PC" = 18 # diamond
)

# Step 3.3: build the graph

# creation of the plot
richness_plot2 <- ggplot(data_2, # requires long-form data
                         aes(x = Richness, y = Point,
                             color = Protocol, shape = Protocol)) +
  
  # add in lines for each location
  geom_segment(
    data = segment_lines2,
    aes(x = Richness, xend = next_richness,
        y = Point, yend = Point),
    inherit.aes = FALSE,
    color = "gray60",
    linewidth = 1
  ) +
  
  # add in all data points
  geom_point(aes(x = Richness),
             size = 5) +

  # group the data by site (y) and by year (x)
  ggh4x::facet_nested(
    Route ~ Year,
    scales = "free_y",
    space = "free_y",
    switch = "y"
  ) +
  
  # reorder the legend and give the points custom colors and shapes
  scale_color_manual(
    values = Protocol_colors,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count combined")
  ) +
  scale_shape_manual(
    values = Protocol_shapes,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count combined")
  ) +
  
  # theme adjustments
  theme_minimal(base_size = 14) + # can change this back to 16 after more troubleshooting
  theme(
    strip.text.y.left = element_text(angle = 0, face = "bold"), # makes Route labels horizontal
    strip.placement = "outside",
    strip.text = element_text(face = "bold"), # makes site, year labels bold
    axis.text.y = element_blank(), # first line to remove the individual Point labels
    axis.ticks.y = element_blank(), # second line to remove the individual Point labels
    panel.spacing = unit(1.2, "lines"),
    panel.background = element_rect(fill = "grey97", color = NA), # add in dark background within each site/year
    legend.position = "top" # if you want the legend at the top, otherwise it'll default to the right.
  ) +
  labs(y = NULL, x = "Species Detected") # remove "Point" from appearing on the left

# Step 3.4: Run the graph to inspect for any errors
#richness_plot2

# Step 4: combine the graphs ==============================================================================================
# create object that combines the two graphs
all_richness_plot <- richness_plot1 + richness_plot2 + 
  plot_layout(
    ncol = 2,
    guides = "collect",
    axis_titles = "collect"
  ) &
  theme(legend.position = "top")

# call the object
all_richness_plot

#ggsave(here::here("results", "richness_per_point_year.png"),
   #    plot = all_richness_plot,
   #    width = 14, height = 10, dpi = 300, units = "in")
```





# Lollipop graph: ARUs vs. point counts, total detections (no sites; all one line)
```{r}
# create a pointYear column, which fixes the lack of a facet wrap grouping the right points =================================================================================================
all_data_plot <- all_data_plot %>%
  mutate(pointYear = paste0(Point, Year))

# manually mutate the data to allow for a better jitter ==============================================================================================
all_data_plot <- all_data_plot %>%
  dplyr::mutate(
    Protocol = factor(Protocol, levels = c("PC", "ARU", "ARU + PC")),
    protocol_order = as.integer(Protocol)
  ) %>%
  dplyr::group_by(Point, pointYear, Richness) %>%   # replace pointID if needed
  dplyr::mutate(
    n_tied = dplyr::n(),
    tie_rank = rank(protocol_order, ties.method = "first"),
    jitter_offset = dplyr::if_else(
      n_tied > 1,
      (tie_rank - (n_tied + 1) / 2) * 0.1,
      0
    ),
    richness_adjusted = Richness + jitter_offset
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-n_tied, -tie_rank, -jitter_offset, -protocol_order)

# final data tweaks before graphing ==============================================================================================

# create dataset for segment lines
segment_lines <- all_data_plot %>%
  dplyr::arrange(Route, Year, Point, richness_adjusted) %>%
  dplyr::group_by(Route, Year, Point) %>%
  dplyr::mutate(next_richness = lead(richness_adjusted)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(!is.na(next_richness))

# order the pointYear column so the graph is easier to interpret
pointYear_order <- all_data_plot %>%
  dplyr::group_by(.data$pointYear) %>%
  dplyr::summarise(
    sort_value = min(.data$richness_adjusted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::arrange(dplyr::desc(.data$sort_value)) %>%
  dplyr::pull("pointYear")

all_data_plot <- all_data_plot %>%
  dplyr::mutate(pointYear = factor(pointYear, levels = pointYear_order))

segment_lines <- segment_lines %>%
  dplyr::mutate(pointYear = factor(pointYear, levels = pointYear_order))



# define graph aesthetics ==============================================================================================
Protocol_colors <- c(
  "ARU" = "#F8766D",
  "PC"  = "#00BFC4",
  "ARU + PC" = "#6F3959"
)

Protocol_shapes <- c(
  "ARU" = 18,
  "PC"  = 18,
  "ARU + PC" = 16 # diamond
)


# create the graph! ==============================================================================================
long_richness_plot <- ggplot(
  all_data_plot,
  aes(x = richness_adjusted, y = pointYear, # x is adjusted for jitter, y is adjusted for discrete point-years
      color = Protocol, shape = Protocol)
) +
  
  # add the lines
  geom_segment( 
    data = segment_lines,
    aes(x = richness_adjusted, xend = next_richness,
        y = pointYear, yend = pointYear),
    inherit.aes = FALSE,
    color = "gray60",
    linewidth = 0.7
  ) +
  
  # add the points
  geom_point(size = 5) +
  scale_color_manual(
    values = Protocol_colors,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count Total")
  ) +
  scale_shape_manual(
    values = Protocol_shapes,
    name = "Detection Method",
    breaks = c("PC", "ARU", "ARU + PC"),
    labels = c("Point Count only", "ARU only", "ARU + Point Count Total")
  ) +
  scale_x_continuous(breaks = scales::breaks_width(2)) + # each vertical line is a whole number (without, the lines are every 1.5)
  theme_minimal(base_size = 16) +
  theme(
    axis.text.y = element_blank(), # y axis is blank since point-years are not important to identify
    legend.position = "bottom" # way too much blank space when it's to the side
  ) +
  labs(y = "Unique Point-Year", x = "Species Detected")

# generate the plot
long_richness_plot






#ggsave(here::here("results", "richness_per_point_year_no_sites.png"),
#      plot = long_richness_plot,
 #     width = 10, height = 14, dpi = 300, units = "in")
```


































# Detectability by species and point-year (lollipop graph)
```{r}
##################################################### 
# 5 Lollipop: Detectability by point-year and species
#####################################################

# data preparation ==============================================================================================

# delete the misc species irrelevant to our study
pc_aru_species <- all_pc_aru_IMPORT %>%
  filter(!CommonName %in% c("----", "Trumpeter Swan", "Sandhill Crane", "Snowy Egret", "Forster's Tern", "Sedge Wren", "Common Tern", "Wilson's Snipe"))

# create a pointYear column (I like to have these be one item)
pc_aru_species <- pc_aru_species %>%
  mutate(pointYear = paste0(Point, Year))

# remove all instances of a non-observation
pc_aru_species_detections <- subset(pc_aru_species, IndivCount > 0)

# drop rows where the same species was observed in the same point-year under the same protocol
# (e.g., dropping all but one Sora per protocol per site per year)
speciesUnique <- pc_aru_species_detections %>%
  distinct(pointYear, Protocol, CommonName)

# pivot to wide:
# for each species at each site, specify which protocol found it
speciesUnique_wide <- speciesUnique %>%
  dplyr::mutate(detected = 1) %>%
  pivot_wider(names_from = Protocol, values_from = detected, values_fill = 0) %>%
  dplyr::mutate(Detection_Category = case_when(
    ARU == 1 & PC == 0 ~ "Only ARU",
    ARU == 0 & PC == 1 ~ "Only PC",
    ARU == 1 & PC == 1 ~ "ARU + PC",
    TRUE ~ "Neither"
  ))

# create list of species wholly missed by a detection category
# and the number of times a given detection category did find them
missed_species <- speciesUnique_wide %>%
    dplyr::group_by(CommonName, Detection_Category) %>%
    dplyr::summarise(count = n(), .groups = "drop") %>%
    filter(Detection_Category != "Neither")

# Define the order of categories for consistent positioning
category_levels <- c("Only ARU", "Only PC", "ARU + PC")
missed_species <- missed_species %>%
  dplyr::mutate(Detection_Category = factor(Detection_Category, levels = category_levels))

# Reorder species by total detections
species_order <- missed_species %>%
  dplyr::group_by(CommonName) %>%
  dplyr::summarise(total = sum(count)) %>%
  dplyr::arrange(total) %>%
  pull(CommonName)

# make sure missed_species is understood as factors 
missed_species <- missed_species %>%
  dplyr::mutate(CommonName = factor(CommonName, levels = species_order))

# graph components ==============================================================================================

# change x variable: percent of points with the detection
# (there are 60 point-years)
missed_species$count_per <- (missed_species$count /60)*100


# step 0: build the dataset for the line segments of the graph
segmentsSpecies <- missed_species %>%
  group_by(CommonName) %>%
arrange(Detection_Category) %>%
  dplyr::reframe(
    x_start = count_per[-length(count_per)],
    x_end   = count_per[-1],
    y       = CommonName[1]
  )

# Step 1: Calculate max n per CommonName and get corresponding Detection_Category
max_n_per_common <- missed_species %>%
  dplyr::group_by(CommonName) %>%
  dplyr::mutate(max_n = max(count_per)) %>%
  filter(count_per == max_n) %>%
  slice(1) %>%  # in case of ties
  ungroup() %>%
  select(CommonName, Detection_Category_max = Detection_Category, max_n)

# Step 2: Identify CommonNames that appear only with "Only ARU"
only_aru_common <- missed_species %>%
  dplyr::group_by(CommonName) %>%
  dplyr::summarize(only_aru = all(Detection_Category == "Only ARU")) %>%
  ungroup()

# Step 3: Merge and assign custom order
ordering_info <- max_n_per_common %>%
  dplyr::left_join(only_aru_common, by = "CommonName") %>%
  dplyr::mutate(order_group = case_when(
    only_aru ~ 1,
    Detection_Category_max == "Only ARU" ~ 2,
    Detection_Category_max == "ARU + PC" ~ 3,
    TRUE ~ 4  # optional, in case of unexpected categories
  ))

# Step 4: Join back to original dataframe and arrange
species_miss_counts_ordered <- missed_species %>%
  left_join(ordering_info, by = "CommonName") %>%
  arrange(order_group, max_n, CommonName)

# re-order
row_id <- seq(1, nrow(species_miss_counts_ordered))
species_miss_counts_ordered$row_id = row_id
species_miss_counts_ordered <- species_miss_counts_ordered %>%
  arrange(desc(row_id))
species_order <- unique(species_miss_counts_ordered$CommonName)

# build the plot! ==============================================================================================

# Plot
reverse_detplot <- ggplot(missed_species,
                          aes(x = count_per, y = CommonName,
                              color = Detection_Category,
                              shape = Detection_Category)) +
  geom_segment(data = segmentsSpecies,
               aes(x = x_start, xend = x_end, y = y, yend = y),
               inherit.aes = FALSE,
               color = "grey50", size = 1) +
  geom_point(size = 5) +
  scale_color_manual(values = c(
    "Only ARU" = "coral1",
    "Only PC" = "dodgerblue",
    "ARU + PC" = "goldenrod1"),
    name = "Detected by:",
    breaks = c("Only ARU", "Only PC", "ARU + PC"),
    labels = c("Only ARU", "Only PC", "Both Methods")
  ) +
  scale_shape_manual(values = c(
    "Only ARU" = 16,
    "Only PC" = 16,
    "ARU + PC" = 16),
    name = "Detected by:",
    breaks = c("Only ARU", "Only PC", "ARU + PC"),
    labels = c("Only ARU", "Only PC", "Both Methods")
  ) +
  labs(
    #title = "Species Detection by Survey Protocol",
    x = "Percent of Point-Years With Detection",
    y = "Species",
    color = "Detected by:", 
    shape = "Detected by:"
  ) +
  scale_y_discrete(limits = species_order) +
  theme_minimal(base_size = 14)

# run the plot
reverse_detplot

# inspect for errors!

# Export image to GitHub ==============================================================================================

#ggsave(here::here("results", "ARU_v_PC_species.png"),
 #     plot = reverse_detplot,
 #     width = 7, height = 9, dpi = 300, units = "in")
```


# Did ARUs find species that have never been detected at a site?
```{r}
# import the historical dataset ==============================================================================================
historic_pc_IMPORT <- readr::read_csv(here::here("CSVs/point_counts", "historical_pc_data.csv"))

# remove the species that aren't relevant to this study
historic_pc_data <- historic_pc_IMPORT %>%
 filter(!CommonName %in% c("----", "Trumpeter Swan", "Sandhill Crane", "Snowy Egret", "Forster's Tern", "Sedge Wren", "Common Tern", "Wilson's Snipe")) 

# remove the sites not shared with our ARU data
historic_pc_data <- historic_pc_data %>%
  filter(!TransectName %in% c("DUNC", "STIG"))

# rename TransectName to Site
historic_pc_data <- historic_pc_data %>% 
  dplyr::rename(Site = TransectName)

# inspect, summarize the historical data ==============================================================================================
# list all species found for each site at some point
# and add the number of times it was detected over time
historic_pc_summary <- ddply(historic_pc_data, c("Site", "BirdCd"),
                             summarise,
                             no_detections = length(BirdCd),
                             dataset = "historic_pc")

# all sites in historic data set, plus:
## the number of years in the record
## the first year of sampling
historic_sites <- ddply(historic_pc_data, c("Site"),
                        summarise,
                        first_year = min(year),
                        last_year = max(year),
                        time_span = last_year - first_year,
                        pc_years_monitored = length(unique(year)),
                        years_not_monitored = abs(pc_years_monitored - time_span - 1) # the -1 is because the first year counts
                        )
# CONCLUSION: only PETE and CHEB are missing any monitoring years, and each is only missing 1.

# prepare the ARU-only data ==============================================================================================

# we have the data already loaded, so we just need to subset for the ARUs.
ARU_detections <- subset(all_pc_aru_IMPORT, Protocol == "ARU")

# These data are zero-filled; need to remove the zeroes!
ARU_detections <- subset(ARU_detections, IndivCount > 0)

# summarize the ARU data to get a list of all species found per site
ARU_species_summary <- ddply(ARU_detections, c("Route", "BirdCd"),
                             summarise,
                             no_detections = length(BirdCd),
                             dataset = "ARU")


# Combine the ARU and historical PC summary datasets and compare them ==============================================================================================

# ensure the datasets have the same site name
ARU_species_summary <- ARU_species_summary %>%
  dplyr::rename(Site = Route)

# merge the datasets
historic_and_ARUs <- rbind(historic_pc_summary, ARU_species_summary)

# determine if species were missed at a site over time
mismatches <- historic_and_ARUs %>%
  distinct(Site, BirdCd, dataset) %>%
  mutate(present = TRUE) %>%
  pivot_wider(
    names_from = dataset,
    values_from = present,
    values_fill = FALSE
  ) %>%
  filter(historic_pc != ARU) %>%
  mutate(
    Found_In = case_when(
      historic_pc == TRUE & ARU == FALSE ~ "historic only",
      historic_pc == FALSE & ARU == TRUE ~ "ARU only"
    )
  )

# Add columns that show years in historical record and years of ARU monitoring ==============================================================================================

# figure out how many years of ARU data were available per site
ARU_detections$Route <- as.factor(ARU_detections$Route)
ARU_years <- ddply(ARU_detections, c("Route"),
                        summarise,
                        first_year = min(Year),
                        last_year = max(Year),
                        time_span = last_year - first_year,
                        ARU_years_monitored = length(unique(Year)),
                        years_not_monitored = abs(ARU_years_monitored - time_span - 1) # the -1 is because the first year counts
                        )
ARU_years <- ARU_years %>%
    dplyr::rename(Site = Route)


# add those years to the "mismatch" dataset
mismatches <- mismatches %>%
  left_join(
    ARU_years %>%
      select(Site, ARU_years_monitored) %>%
      distinct(),
    by = "Site"
  )

# add the amount of PC monitoring years from a prior dataset
mismatches <- mismatches %>%
  left_join(
    historic_sites %>%
      select(Site, pc_years_monitored) %>%
      distinct(),
    by = "Site"
  )

# which birds were most commonly missing? ==============================================================================================
missingBirds <- mismatches %>%
  dplyr::count(Found_In, BirdCd, name = "times_missed")


# Which birds did ARUs miss in the same years where PCs found them?
# ==============================================================================================

# get a summary dataset of all positive detections from both ARUs and PCs
pc_aru_dataset <- all_pc_aru_IMPORT %>%
 filter(!CommonName %in% c("----", "Trumpeter Swan", "Sandhill Crane", "Snowy Egret", "Forster's Tern", "Sedge Wren", "Common Tern", "Wilson's Snipe")) %>% # had to add the sedge wren and common tern
  dplyr::filter(IndivCount >=1) %>%
  dplyr::group_by(Protocol, Route, Year, Point)

# list all species found at each site in each year for each method
ARU_PC_species_summary <- ddply(pc_aru_dataset, c("Route", "Protocol", "Year", "BirdCd"),
                             summarise,
                             no_detections = length(BirdCd))


# Dataset: list species found only by one method at a site
StudyDiffs <- ARU_PC_species_summary %>%
  distinct(Route, Year, BirdCd, Protocol) %>%   # removes duplicate records
  group_by(Route, Year, BirdCd) %>%
  filter(n_distinct(Protocol) == 1) %>%         # keeps cases with only ARU or only PC
  mutate(
    Positive = case_when(
      Protocol == "ARU" ~ "ARU only",
      Protocol == "PC"  ~ "PC only"
    )
  ) %>%
  ungroup() %>%
  select(Route, Year, BirdCd, Positive)


# summary: number of times each method 
StudyDiffSummaries <- StudyDiffs %>%
  dplyr::count(Positive, name = "ARU_only")









# export results to CSVs ==============================================================================================

# overall comparison with historical record
write.csv(mismatches, here::here("results", "comparison_with_historical_record.csv"))

# table of birds never found at a site but found in ARUs
write.csv(missingBirds, here::here("results", "counts_of_missing_species_in_historic_pc.csv"))

```

# Species richness accumulation curves by validation attempt
```{r}
##############################################################
# Species Richness Accumulation Curves by Validations Attempts
##############################################################

# Read in all (raw) ARU validations
#all_scores <- readr::read_csv(here::here("CSVs/analysis", "ARU_Validated_not_prepped_for_occ.csv"))
all_scores <- read.csv("E:/Documents/Audubon/AGL/ARU/Validation/Verified/AGLARU_Validated_2_27_26.csv") # You can remove this when you run it and use the one above

# Richness accumulation curves with 5 "Validation Attempts"  ==============================================================================================

# Filter to sites that were checked up to 5 rows
test_scores5 <- all_scores %>%
  # keep only distinct rows
  distinct(Route, Point, Date, Year, StartTime, CommonName, Confidence, .keep_all = TRUE) %>%
  # filter to test sites
  filter(
    Year == 2024 &
    (
      Route %in% c("PETE", "BIGM") |
      (Route == "CMBM" & Point %in% c("CMBM122", "CMBM124", "CMBM142", "CMBM144", "CMBM146"))
    )
  ) %>%
  # make sure we are only including the threshold of 5 detections
  group_by(Date, Point, CommonName) %>%
  slice_max(order_by = Confidence, n = 5, with_ties = FALSE) %>%
  ungroup() 

# Report the number of point-years evaluated (n = 14)
test_scores5 %>%
  distinct(Point, Year) 

# Add detection attempts to dataset
test_scores5_attempts <- test_scores5 %>%
  dplyr::arrange(Point, Date, Year, CommonName, Confidence) %>%  
  dplyr::group_by(Point, Date, Year, CommonName) %>%
  dplyr::mutate(det_attempt = row_number()) %>%
  dplyr::ungroup()

# assign validation attempt thresholds
thresholds_5 <- 1:5

# map out point-year level richness by k (number of validation attempts per species/point/date)
richness_by_k_5 <- map_dfr(thresholds_5, function(k) {
  
  test_scores5_attempts %>%
    filter(det_attempt <= k, IndivCount == 1) %>%
    
    # collapse to unique detections per species per point-year
    distinct(Route, Point, Year, CommonName) %>%
    
    # richness per point-year
     dplyr::group_by(Route, Point, Year) %>%
     dplyr::summarize(richness = n_distinct(CommonName), .groups = "drop") %>%
    
    mutate(k = k)
})

# estimate the "gain" in species richness for each validation attempt threshold (1-5)
gain_curve_5 <- richness_by_k_5 %>%
  pivot_wider(names_from = k, values_from = richness) %>%
  dplyr::mutate(
    gain_1 = `1`,
    gain_2 = `2` - `1`,
    gain_3 = `3` - `2`,
    gain_4 = `4` - `3`,
    gain_5 = `5` - `4`
  ) %>%
  dplyr::select(Point, Year, starts_with("gain_")) %>%
  pivot_longer(
    cols = starts_with("gain_"),
    names_to = "step",
    values_to = "new_species"
  ) %>%
  dplyr::mutate(k = case_when(
    step == "gain_1" ~ 1,
    step == "gain_2" ~ 2,
    step == "gain_3" ~ 3,
    step == "gain_4" ~ 4,
    step == "gain_5" ~ 5
  ))

# calculate cumulative richness based on added species
cumulative_curve_5 <- gain_curve_5 %>%
  dplyr::group_by(Point, Year) %>%
  dplyr::arrange(k) %>%
  dplyr::mutate(cumulative_richness = cumsum(new_species)) %>%
  dplyr::ungroup()

# summarize average richness accumulation 
summary_curve_5 <- cumulative_curve_5 %>%
  dplyr::group_by(k) %>%
  dplyr::summarize(
    mean_richness = mean(cumulative_richness),
    sd_richness = sd(cumulative_richness),
    .groups = "drop"
  )

# calculate max richness
max_rich_5 <- summary_curve_5 %>%
  filter(k == 5) %>%
  pull(mean_richness)

# plot accumulation curve
rich_accumulation_5attempts <- summary_curve_5 %>%
  dplyr::mutate(prop_max = mean_richness / max_rich_5) %>%
  ggplot(aes(x = k, y = prop_max)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 4, linetype = "dashed") +
  labs(
    x = "Validated Detections per Species/ARU/Day",
    y = "Cumulative Proportion of Total Richness",
    title = "Saturation of Species Detections with Validation Effort"
  ) +
  theme_classic()

# Species rate of discovery test with 5 "Validation Attempts"  ==============================================================================================
# Rate of discovery by individual species: what number of detections/day/species were species first detected positive?

# summarize detections for each species by attempt number
det_by_species_5 <- test_scores5_attempts %>%
  filter(IndivCount == 1) %>%
  group_by(CommonName, Point, Date, det_attempt) %>%
  slice_max(order_by = Confidence, n = 1, with_ties = FALSE) %>%
  distinct(CommonName, Point, Date, det_attempt) %>%
  dplyr::group_by(CommonName, det_attempt) %>%
  dplyr::summarize(total_detected = n(), .groups = "drop")

# calculate the proportion out of total detected for each attempt number
det_prop_5 <- det_by_species_5 %>%
  dplyr::group_by(CommonName) %>%
  dplyr::arrange(det_attempt) %>%
  dplyr::mutate(
    prop_total = total_detected / sum(total_detected),
    cum_prop = cumsum(prop_total)
  ) %>%
  ungroup()

# plot cumulative detection curve
sp_det_curve_5attempts <- ggplot(det_prop_5, aes(x = det_attempt, y = cum_prop)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ CommonName) +
  labs(
    x = "Validation Attempt",
    y = "Cumulative Proportion of Total Positive Detections",
    title = "Species-level Accumulation of Detections with Validation Effort"
  ) +
  theme_classic()

# Richness accumulation curves with all data (4 Validation Attempts)  ==============================================================================================

# All Sites
test_scores_all <- all_scores %>%
  distinct(Route, Point, Date, Year, StartTime, CommonName, Confidence, .keep_all = TRUE) %>%
  group_by(Date, Point, CommonName) %>%
  # make sure we are only including the threshold of 4 detections
  slice_max(order_by = Confidence, n = 4, with_ties = FALSE) %>%
  ungroup() 

# NOTE: number of point-years evaluated (n = 60)
test_scores_all %>%
  distinct(Point, Year) 

# Add detection attempts to dataset
test_scores_all_attempts <- test_scores_all %>%
  dplyr::arrange(Point, Date, Year, CommonName, Confidence) %>%  
  dplyr::group_by(Point, Date, Year, CommonName) %>%
  dplyr::mutate(det_attempt = row_number()) %>%
  dplyr::ungroup()

# assign validation attempt thresholds
thresholds_all <- 1:4

# map out point-year level richness by k (number of validation attempts per species/point/date)
richness_by_k_all <- map_dfr(thresholds_all, function(k) {
  
  test_scores_all_attempts %>%
    filter(det_attempt <= k, IndivCount == 1) %>%
    
    # collapse to unique detections per species per point-year
    distinct(Route, Point, Year, CommonName) %>%
    
    # richness per point-year
     dplyr::group_by(Route, Point, Year) %>%
     dplyr::summarize(richness = n_distinct(CommonName), .groups = "drop") %>%
    
    mutate(k = k)
})

# estimate the "gain" in species richness for each validation attempt threshold (1-5)
gain_curve_all <- richness_by_k_all %>%
  pivot_wider(names_from = k, values_from = richness) %>%
  dplyr::mutate(
    gain_1 = `1`,
    gain_2 = `2` - `1`,
    gain_3 = `3` - `2`,
    gain_4 = `4` - `3`
  ) %>%
  dplyr::select(Point, Year, starts_with("gain_")) %>%
  pivot_longer(
    cols = starts_with("gain_"),
    names_to = "step",
    values_to = "new_species"
  ) %>%
  dplyr::mutate(k = case_when(
    step == "gain_1" ~ 1,
    step == "gain_2" ~ 2,
    step == "gain_3" ~ 3,
    step == "gain_4" ~ 4
  ))

# calculate cumulative richness based on added species
cumulative_curve_all <- gain_curve_all %>%
  dplyr::group_by(Point, Year) %>%
  dplyr::arrange(k) %>%
  dplyr::mutate(cumulative_richness = cumsum(new_species)) %>%
  dplyr::ungroup()

# summarize average richness accumulation 
summary_curve_all <- cumulative_curve_all %>%
  dplyr::group_by(k) %>%
  dplyr::summarize(
    mean_richness = mean(cumulative_richness),
    sd_richness = sd(cumulative_richness),
    .groups = "drop"
  )

# calculate max richness
max_rich_all <- summary_curve_all %>%
  filter(k == 4) %>%
  pull(mean_richness)

# plot accumulation curve
rich_accumulation_all <- summary_curve_all %>%
  dplyr::mutate(prop_max = mean_richness / max_rich_all) %>%
  ggplot(aes(x = k, y = prop_max)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Validated Detections per Species/ARU/Day",
    y = "Cumulative Proportion of Total Richness",
    title = "Saturation of Species Detections with Validation Effort"
  ) +
  theme_classic()

# Species rate of discovery with all data (4 Validation Attempts) ==============================================================================================
# Rate of discovery by individual species: what number of detections/day/species were species first detected positive?

# summarize detections for each species by attempt number
det_by_species_all <- test_scores_all_attempts %>%
  filter(IndivCount == 1) %>%
  group_by(CommonName, Point, Date, det_attempt) %>%
  slice_max(order_by = Confidence, n = 1, with_ties = FALSE) %>%
  distinct(CommonName, Point, Date, det_attempt) %>%
  dplyr::group_by(CommonName, det_attempt) %>%
  dplyr::summarize(total_detected = n(), .groups = "drop")

# calculate the proportion out of total detected for each attempt number
det_prop_all <- det_by_species_all %>%
  dplyr::group_by(CommonName) %>%
  dplyr::arrange(det_attempt) %>%
  dplyr::mutate(
    prop_total = total_detected / sum(total_detected),
    cum_prop = cumsum(prop_total)
  ) %>%
  ungroup()

# plot cumulative detection curve
sp_det_curve_all <- ggplot(det_prop_all, aes(x = det_attempt, y = cum_prop)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ CommonName) +
  labs(
    x = "Validation Attempt",
    y = "Cumulative Proportion of Total Positive Detections",
    title = "Species-level Accumulation of Detections with Validation Effort"
  ) +
  theme_classic()
```

# BirdNET precision for species-specific confidence values
```{r}
##############################################################
# BirdNET Precision for Species-Specific Confidence Values
##############################################################

# Prep dataset
all_scores <- all_scores %>%
  distinct(Route, Point, Date, Year, StartTime, CommonName, Confidence, .keep_all = TRUE)

# Confidence Histograms ==============================================================================================
# Plot validated confidence histograms for each species
species_conf_hist <- ggplot(all_scores, aes(x = Confidence)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "skyblue") +
  facet_wrap(~ CommonName) +
  theme_classic()

# Precision Curves ==============================================================================================
# Precision = True Positives / (True + False Positives)

# Create a clean dataframe, keeping only the highest TP per species/ARU/day, otherwise all FP
all_scores_clean <- all_scores %>%
  distinct(Route, Point, Date, Year, StartTime, CommonName, Confidence, .keep_all = TRUE) %>%
  group_by(CommonName, Date, Point) %>%
  
  # if any true positive exists
  group_modify(~{
    
    if(any(.x$IndivCount == 1)) {
      
      # keep highest-confidence TP only
      .x %>%
        filter(IndivCount == 1) %>%
        slice_max(Confidence, n = 1, with_ties = FALSE)
      
    } else {
      
      # otherwise keep all FP
      .x
    }
  }) %>%
  ungroup()

# Create confidence bins to calculate precision within each
df_bins <- all_scores_clean %>%
  
  mutate(
    Confidence_Bin = cut(
      Confidence,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  )

precision_bins <- df_bins %>%
  
  dplyr::group_by(CommonName, Confidence_Bin) %>%
  
  dplyr::summarise(
    TP = sum(IndivCount == 1),
    FP = sum(IndivCount == 0),
    Precision = TP / (TP + FP),
    N = n(),
    .groups = "drop"
  )

# Plot precision by species and confidence bin
sp_conf_precision <- ggplot(precision_bins,
       aes(x = Confidence_Bin,
           y = Precision,
           group = 1)) +
  
  geom_line() +
  geom_point() +
  
  facet_wrap(~CommonName) +
  
  ylim(0, 1) +
  
  labs(
    x = "Confidence bin",
    y = "Precision"
  ) +
  
  theme_classic() +
  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```



















